Localization properties, observable character and computational reliability of the linear response in open systems

### 2018-06-26 AnalyticStructure_main.tex
update

\subsection{Long range behavior of the eigenvalue solution of the Schroedinger equation}
\label{SEopenSystem}
Consider a (one-body) wavefunction $\ket{\psi}$ solution of the equation
$\op H\ket{\psi}=\eps\ket{\psi}$, where we split the Hamiltonian operator into the sum of
a kinetic $\op T$ and a potential $\op V \equiv \op H - \op T$ term.
In view of analysing the impact of the BC, it is instructive to transform 
the Schroedinger Equation in a \emph{scattering} problem, by writing
\be\lb{SEHelmholtz1}
\ket{\psi} = \op G_H(\eps)\ket{\op V\psi}\;,
\ee
where we have introduced the Green function of the Helmholtz operator $\op G_H(\eps) = (\eps-\op T)^{-1}$,
resolvent of the free-particle hamltonian $H_s \equiv \op T$.

@@

It is easy to see from these concepts that a computational treatment that is tailored for the discretization of bound states
will not be effective for continuum states, due to the different long-range asymptotic behaviour - which in turn is a clear consequence of the 
BC of the problem. In a numerical treatment, discretizing the problem with a \emph{finite}
number of degrees of freedom, we will only have acces to \emph{pseudo-} continuum states,
whose energy values and density of states will depend on the representation of the kinetic operator
in the basis set. As a consequence, numerical eigenstates with $\epsilon > 0$ will have a value of their energy that will \emph{always}
depend on the computational setup.

@@

Such locality property of the fluctuation state can be directly confirmed for static perturbations at $\omega=0$.
Let us consider a static perturbing operator $\op\Phi_s$ that can be restricted to a finite domain. 
Any local operator, like e.g. a static electric field in the direction $\mathbf F$, $\Phi_s(\r) = -\mathbf F \cdot \r$, would satisfy such property.
We consider the perturbed hamiltonian $\hnot[\rho_{\Phi_s}] + \Phi_s$, where we indicate with $\rho_{\Phi_s}$ the GS density of the
perturbed system. We assume that the perturbation strength ($F$ in the above exemple) is sufficiently small to be consistent with a LR 
treatment of the problem. 
In this case, the occupied eigenfunction $\ket{\psi_p^{\Phi_s}}$ of the perturbed hamiltonian 
can be extracted with traditional GS techniques, as by the hypothesis of Eq.~\eqref{SEHelmholtz1},
they would exhibit bound-state behaviour.\footnote{The fulfilling of Eq.~\eqref{SEHelmholtz1} for energies below threshold
is not \textit{per se} a guarantee of a localized $\ket{\psi}$. With such equation we demonstrate that, \emph{in the case when} a localized
solution exists, this cannot be found for energies above threshold.}
\begin{figure}[t]
\includegraphics[scale=0.56]{CO_alpha+D0_vsBoxSize.eps}
\caption{\label{co_alphaStatic}(Color online) Convergence of the statical polarizability of the $CO$ with respect to the size of the computational domain.}
\end{figure}

From the linearized Schroedinger Equation of the perturbed problem, it is easy to see that the fluctuation state at $\omega=0$ can be expressed
in terms of the corresponding bound state of the perturbed Hamiltonian, i.e.
$\ket{f_p^{\Phi_s}(0)} \simeq \op Q_0 \ket{\psi_p^{\Phi_s}}$.
For any value of $p$, the locality of the $\ket{f_p^{\Phi_s}(0)}$ therefore directly stems from the
bound-state behaviour of $\ket{\psi_p^{\Phi_s}}$. As $\omega=0$ is by hypothesis 
lower than $|\eps_p|$, this is consistent with the above considerations.

@@

The analysis described above has shown that the building blocks of the response density $\{\ket{f_p(\omega)}\}$ 
are susceptible to exhibit bound-state long-range behaviour \emph{only} when $\omega < |\eps_p|$, 
while they behave as unbound states for values of $\omega$ above the threshold levels.

This fact has interesting implication in the evaluation of LR quantities.
%
% In the next section we will discuss the implication of the existence of these
% two distinct regimes on the numerical convergence of the observables computed within the LR framework. 
% Before doing this we propose an analogous analysis with the aim to understand the localization 
% properties of the excitation-based description of the LR.
% 
% The localization properties of fluctuation states allows us to assess the reliability of the linear response functional
% computed using the given basis set. 
%
Indeed, writing the response 
density in terms of fluctuation states implies that 
\be\lb{LinearResponseFunctDef2}
\braket{\delta\op O}_\Phi(\omega) = 2\mathrm{Re} \left(\sum_p \bra{\psi_p}\op O\ket{f_p^\Phi(\omega)}\right) \;.
\ee
%so the computation of the linear response functional passes through the evaluation of the scalar products indicated in \eqref{LinearResponseFunctDef2}.
Let us suppose that also the observable of interest $\op O$ exhibit the locality properties of the potentials $ \op V$, $\op \Phi$, $\op V'$.
Once again, it would be enough to consider a local operator for this condition to be met.
We observe that there exists two 
distinct regimes, on the basis of value of $\omega$. 
The first one, which we will refer to as the \emph{below threshold} regime, is realized when
$\omega$ is lower than the ionization potential $|\eps_h|$.
In this case all the fluctuation states are below their threshold level and equation \eqref{LinearResponseFunctDef2} 
is expressed in terms of genuinely localized 
quantities. The quantities $\braket{\psi_p \op O| f_p^\Phi(\omega)}$, 
being scalr products of bound-state-like objects, may thus be evaluated with computational setup that
are similar to the ones usually employed for GS calculations.
This consideration proves why, when a given computer code employs established GS numerical techniques, 
\emph{it is much easier to converge LR quantities for $\omega$ below threshold}.
The convergence in this regime is ``strong'', in the sense that the basis functions employed are able to express 
with arbitrary precision \emph{both} the states $\ket{\op O \psi_p}$ and $\ket{f_p^\Phi(\omega)}$.
This is of course true even when the above states are only implicitly expressed by the employed LR treatment,
 as e.g. in the case of Krylov spaces generated from the states $\ket{\op\Phi(\omega) \psi_p}$ (see.e.g. \cite{baroni2008,linlinKPM}).

@@

On the contrary, the computation of the linear response functional can be much more demanding in the \emph{above threshold regime}, realized for $\omega>|\eps_h|$, in which fluctuation 
states start behaving as unbound wave functions. In this case a computational description of fluctuation states is only meaningful in ``weak'' sense,
\emph{i.e.} inside the scalar products of 
\eqref{LinearResponseFunctDef2}. 
In this case $\ket{f_p^\Phi(\omega)}$ is a unbound, delocalized function, and 
the bound state character of $\{\bra{\psi_p}\op O\}$ behave as a regulator. 
In this regime \emph{only} the scalar products of Eq.~\eqref{LinearResponseFunctDef2} can be evaluated in a localized domain.
This observation indicates that, above threshold, a computational setup 
provides a reliable assessment of \eqref{LinearResponseFunctDef2} only if it is able to express an
unbiased description of the fluctuation states in the support of $\{\bra{\psi_p}\op O\}$.
The non-vanishing spherical-wave behaviour in the long range of $\ket{f_p^\Phi(\omega)}$ is therefore responsible
of this convergence difficulties, especially for high frequencies.

@@

We show that Casida equations are equivalent to the equations of motion for the excited states \eqref{ExcitationOperatorsDef3}. To this scope we introduce an explicit basis $\{\ket{\eta}\}$
in the subspace of empty states so that both $\ket{\phi^a_p}$ and $\bra{\chi^a_p}$ can be represented as
\be
\ket{\phi^a_p} = \sum_\eta X^a_{p\eta}\ket{\eta} \;, \q
\bra{\chi^a_p} = \sum_\eta}\bra{\eta}Y^a_{p\eta} \;.
\ee
Plugging this expansion into equations \eqref{ExcitationOperatorsDef3} and projecting on a arbitrary element $\ket{\rho}$ of the basis provides a linear systems of equations for the coefficients
$X^a_{p\eta}$ and $Y^a_{p\eta}$, namely
\begin{align}
 \sum_\eta(H_{0\rho\eta}-\eps_p\delta_{\rho\eta})X^a_{p\eta} + \bra{\rho}\op V'[\op E_a]\ket{\psi_p} &= \Omega_a X^a_{p\rho} \;, \nn \\
 -\sum_\eta (H_{0\eta\rho}-\eps_p\delta_{\rho\eta})Y^a_{p\eta} - \bra{\psi_p}\op V'[\op E_a]\ket{\rho}& = \Omega_a Y^a_{p\rho} \;. \nn
\end{align}
Using this basis, excitation operators are expressed as a linear combination of transition operators $\excite{p}{\eta} \equiv \ketbra{\eta}{\psi_p}$ and $\decay{\eta}{p} \equiv \ketbra{\psi_p}{\eta}$,
as follows
\be\lb{ExcitationOpBasisTransition1}
\op E_a = \sum_{p\eta}\excite{p}{\eta}X^a_{p\eta}+\decay{\eta}{p}Y^a_{p\eta} \;.
\ee
The contribution of the coupling operator in the equations of motion of excited states can expressed by means of the coupling kernel \eqref{CouplingKernelDef1} as
\begin{align}
& \bra{\rho}\op V'[\op E_a]\ket{\psi_p} = \sum_{q\eta}\left(U[\decay{\rho}{p};\excite{q}{\eta}]X^a_{q\eta} + U[\decay{\rho}{p};\decay{\eta}{q}]Y^a_{q\eta}\right) \;,\nn \\
& \bra{\psi_p}\op V'[\op E_a]\ket{\rho} = \sum_{q\eta}\left(U[\excite{p}{\rho};\excite{q}{\eta}]X^a_{q\eta} + U[\excite{p}{\rho};\decay{\eta}{q}]Y^a_{q\eta}\right) \;, \nn
\end{align}
and so equations \eqref{ExcitationOperatorsDef3} can be recasted in matrix form
\be\lb{ExcitationMatrixEq1}
\sum_{q\eta}\mat{F_{p\rho}^{q\eta} &  D_{p\rho}^{q\eta}  \\
- E_{p\rho}^{q\eta} & - G_{p\rho}^{q\eta} }
\mat{X^a_{q\eta} \\ Y^a_{q\eta}} = \Omega_a \mat{X^a_{p\rho} \\ Y^a_{p\rho}}
\ee
where
\begin{align}
& F_{p\rho}^{q\eta} = (H_{0\rho\eta}-\eps_p\delta_{\rho\eta})\delta_{pq} + U[\decay{\rho}{p};\excite{q}{\eta}] \;,\nn \\
& D_{p\rho}^{q\eta} = U[\decay{\rho}{p};\decay{\eta}{q}] \;,\nn \\
& E_{p\rho}^{q\eta} = U[\excite{p}{\rho};\excite{q}{\eta}] \;,\nn \\
& G_{p\rho}^{q\eta} = (H_{0\eta\rho}-\eps_p\delta_{\rho\eta})\delta_{pq} + U[\excite{p}{\rho};\decay{\eta}{q}] \;. \nn
\end{align}
Finally, exploiting the symmetry properties of both the unperturbed Hamiltonian and of the coupling kernel it follows that $F=G$ and $D=E$ so that equations
\eqref{ExcitationMatrixEq1} reduce to the usual formulation of the Casida equations in the basis of transitions.

>>>>

\subsection{Long range behavior of the eigenvalue solution of the Schroedinger equation}
\label{SEopenSystem}
Consider a (one-body) wavefunction $\ket{\psi}$ solution of the equation
$\op H\ket{\psi}=\eps\ket{\psi}$, where we split the Hamiltonian operator into the sum of
a kinetic $\op T$ and a potential $\op V \equiv \op H - \op T$ term.
In view of analysing the impact of the BC, it is instructive to transform 
the Schroedinger Equation in a \emph{scattering} problem, by writing
\be\lb{SEHelmholtz1}
\ket{\psi} = \op G_H(\eps)\ket{\op V\psi}\;,
\ee
where we have introduced the Green function of the Helmholtz operator $\op G_H(\eps) = (\eps-\op T)^{-1}$,
resolvent of the free-particle Hamiltonian $H_s \equiv \op T$.

@@

It is easy to see from these concepts that a computational treatment that is tailored for the discretization of bound states
will not be effective for continuum states, due to the different long-range asymptotic behaviour - which in turn is a clear consequence of the 
BC of the problem. In a numerical treatment, discretizing the problem with a \emph{finite}
number of degrees of freedom, we will only have access to \emph{pseudo-} continuum states,
whose energy values and density of states will depend on the representation of the kinetic operator
in the basis set. As a consequence, numerical eigenstates with $\epsilon > 0$ will have a value of their energy that will \emph{always}
depend on the computational setup.

@@

Such locality property of the fluctuation state can be directly confirmed for static perturbations at $\omega=0$.
Let us consider a static perturbing operator $\op\Phi_s$ that can be restricted to a finite domain. 
Any local operator, like e.g. a static electric field in the direction $\mathbf F$, $\Phi_s(\r) = -\mathbf F \cdot \r$, would satisfy such property.
We consider the perturbed hamiltonian $\hnot[\rho_{\Phi_s}] + \Phi_s$, where we indicate with $\rho_{\Phi_s}$ the GS density of the
perturbed system. We assume that the perturbation strength ($F$ in the above example) is sufficiently small to be consistent with a LR 
treatment of the problem. 
In this case, the occupied eigenfunction $\ket{\psi_p^{\Phi_s}}$ of the perturbed hamiltonian 
can be extracted with traditional GS techniques, as by the hypothesis of Eq.~\eqref{SEHelmholtz1},
they would exhibit bound-state behavior.\footnote{The fulfilling of Eq.~\eqref{SEHelmholtz1} for energies below threshold
is not \textit{per se} a guarantee of a localized $\ket{\psi}$. With such equation we demonstrate that, \emph{in the case when} a localized
solution exists, this cannot be found for energies above threshold.}
\begin{figure}[t]
\includegraphics[scale=0.56]{CO_alpha+D0_vsBoxSize.eps}
\caption{\label{co_alphaStatic}(Color online) Convergence of the statical polarizability of the $CO$ with respect to the size of the computational domain.}
\end{figure}

From the linearized Schroedinger Equation of the perturbed problem, it is easy to see that the fluctuation state at $\omega=0$ can be expressed
in terms of the corresponding bound state of the perturbed Hamiltonian, i.e.
$\ket{f_p^{\Phi_s}(0)} \simeq \op Q_0 \ket{\psi_p^{\Phi_s}}$.
For any value of $p$, the locality of the $\ket{f_p^{\Phi_s}(0)}$ therefore directly stems from the
bound-state behavior of $\ket{\psi_p^{\Phi_s}}$. As $\omega=0$ is by hypothesis 
lower than $|\eps_p|$, this is consistent with the above considerations.

@@

The analysis described above has shown that the building blocks of the response density $\{\ket{f_p(\omega)}\}$ 
are susceptible to exhibit bound-state long-range behavior \emph{only} when $\omega < |\eps_p|$, 
while they behave as unbound states for values of $\omega$ above the threshold levels.

This fact has interesting implication in the evaluation of LR quantities.
%
% In the next section we will discuss the implication of the existence of these
% two distinct regimes on the numerical convergence of the observables computed within the LR framework. 
% Before doing this we propose an analogous analysis with the aim to understand the localization 
% properties of the excitation-based description of the LR.
% 
% The localization properties of fluctuation states allows us to assess the reliability of the linear response functional
% computed using the given basis set. 
%
Indeed, writing the response 
density in terms of fluctuation states implies that 
\be\lb{LinearResponseFunctDef2}
\braket{\delta\op O}_\Phi(\omega) = 2\mathrm{Re} \left(\sum_p \bra{\psi_p}\op O\ket{f_p^\Phi(\omega)}\right) \;.
\ee
%so the computation of the linear response functional passes through the evaluation of the scalar products indicated in \eqref{LinearResponseFunctDef2}.
Let us suppose that also the observable of interest $\op O$ exhibit the locality properties of the potentials $ \op V$, $\op \Phi$, $\op V'$.
Once again, it would be enough to consider a local operator for this condition to be met.
We observe that there exists two 
distinct regimes, on the basis of value of $\omega$. 
The first one, which we will refer to as the \emph{below threshold} regime, is realized when
$\omega$ is lower than the ionization potential $|\eps_h|$.
In this case all the fluctuation states are below their threshold level and equation \eqref{LinearResponseFunctDef2} 
is expressed in terms of genuinely localized 
quantities. The quantities $\braket{\psi_p \op O| f_p^\Phi(\omega)}$, 
being scalr products of bound-state-like objects, may thus be evaluated with computational setup that
are similar to the ones usually employed for GS calculations.
This consideration proves why, when a given computer code employs established GS numerical techniques, 
\emph{it is much easier to converge LR quantities for $\omega$ below threshold}.
The convergence in this regime is ``strong'', in the sense that the basis functions employed are able to express 
with arbitrary precision \emph{both} the states $\ket{\op O \psi_p}$ and $\ket{f_p^\Phi(\omega)}$.
This is of course true even when the above states are only implicitly expressed by the employed LR treatment,
 as e.g. in the case of Krylov spaces generated from the states $\ket{\op\Phi(\omega) \psi_p}$ \cite{baroni2006,baroni2008,linlinKPM}.

@@

On the contrary, the computation of the linear response functional can be much more demanding in the \emph{above threshold regime}, realized for $\omega>|\eps_h|$, in which fluctuation 
states start behaving as unbound wave functions. In this case a computational description of fluctuation states is only meaningful in ``weak'' sense,
\emph{i.e.} inside the scalar products of 
\eqref{LinearResponseFunctDef2}. 
In this case $\ket{f_p^\Phi(\omega)}$ is a unbound, delocalized function, and 
the bound state character of $\{\bra{\psi_p}\op O\}$ behave as a regulator. 
In this regime \emph{only} the scalar products of Eq.~\eqref{LinearResponseFunctDef2} can be evaluated in a localized domain.
This observation indicates that, above threshold, a computational setup 
provides a reliable assessment of \eqref{LinearResponseFunctDef2} only if it is able to express an
unbiased description of the fluctuation states in the support of $\{\bra{\psi_p}\op O\}$.
The non-vanishing spherical-wave behavior in the long range of $\ket{f_p^\Phi(\omega)}$ is therefore responsible
of this convergence difficulties, especially for high frequencies.

@@

We show that Casida equations are equivalent to the equations of motion for the excited states \eqref{ExcitationOperatorsDef3}. To this scope we introduce an explicit basis $\{\ket{\eta}\}$
in the subspace of empty states so that both $\ket{\phi^a_p}$ and $\bra{\chi^a_p}$ can be represented as
\be
\ket{\phi^a_p} = \sum_\eta X^a_{p\eta}\ket{\eta} \;, \qq
\bra{\chi^a_p} = \sum_\eta\bra{\eta}Y^a_{p\eta} \;.
\ee
Plugging this expansion into equations \eqref{ExcitationOperatorsDef3} and projecting on a arbitrary element $\ket{\rho}$ of the basis provides a linear systems of equations for the coefficients
$X^a_{p\eta}$ and $Y^a_{p\eta}$, namely
\begin{align}
 \sum_\eta(H_{0\rho\eta}-\eps_p\delta_{\rho\eta})X^a_{p\eta} + \bra{\rho}\op V'[\op E_a]\ket{\psi_p} &= \Omega_a X^a_{p\rho} \;, \nn \\
 -\sum_\eta (H_{0\eta\rho}-\eps_p\delta_{\rho\eta})Y^a_{p\eta} - \bra{\psi_p}\op V'[\op E_a]\ket{\rho}& = \Omega_a Y^a_{p\rho} \;. \nn
\end{align}
Using this basis, excitation operators are expressed as a linear combination of transition operators $\excite{p}{\eta} \equiv \ketbra{\eta}{\psi_p}$ and $\decay{\eta}{p} \equiv \ketbra{\psi_p}{\eta}$,
as follows
\be\lb{ExcitationOpBasisTransition1}
\op E_a = \sum_{p\eta}\excite{p}{\eta}X^a_{p\eta}+\decay{\eta}{p}Y^a_{p\eta} \;.
\ee
The contribution of the coupling operator in the equations of motion of excited states can expressed by means of the coupling kernel \eqref{CouplingKernelDef1} as
\begin{align}
& \bra{\rho}\op V'[\op E_a]\ket{\psi_p} = \sum_{q\eta}\left(U[\decay{\rho}{p};\excite{q}{\eta}]X^a_{q\eta} + U[\decay{\rho}{p};\decay{\eta}{q}]Y^a_{q\eta}\right) \;,\nn \\
& \bra{\psi_p}\op V'[\op E_a]\ket{\rho} = \sum_{q\eta}\left(U[\excite{p}{\rho};\excite{q}{\eta}]X^a_{q\eta} + U[\excite{p}{\rho};\decay{\eta}{q}]Y^a_{q\eta}\right) \;, \nn
\end{align}
and so equations \eqref{ExcitationOperatorsDef3} can be recasted in matrix form
\be\lb{ExcitationMatrixEq1}
\sum_{q\eta}\mat{F_{p\rho}^{q\eta} &  D_{p\rho}^{q\eta}  \\
- E_{p\rho}^{q\eta} & - G_{p\rho}^{q\eta} }
\mat{X^a_{q\eta} \\ Y^a_{q\eta}} = \Omega_a \mat{X^a_{p\rho} \\ Y^a_{p\rho}}
\ee
where
\begin{align}
& F_{p\rho}^{q\eta} = (H_{0\rho\eta}-\eps_p\delta_{\rho\eta})\delta_{pq} + U[\decay{\rho}{p};\excite{q}{\eta}] \;,\nn \\
& D_{p\rho}^{q\eta} = U[\decay{\rho}{p};\decay{\eta}{q}] \;,\nn \\
& E_{p\rho}^{q\eta} = U[\excite{p}{\rho};\excite{q}{\eta}] \;,\nn \\
& G_{p\rho}^{q\eta} = (H_{0\eta\rho}-\eps_p\delta_{\rho\eta})\delta_{pq} + U[\excite{p}{\rho};\decay{\eta}{q}] \;. \nn
\end{align}
Finally, exploiting the symmetry properties of both the unperturbed Hamiltonian and of the coupling kernel it follows that $F=G$ and $D=E$ so that equations
\eqref{ExcitationMatrixEq1} reduce to the usual formulation of the Casida equations in the basis of transitions.
